---
title: "Is automatic or manual transmission better for MPG?"
author: "Bankbintje"
date: "17th august 2015"
output: pdf_document
---
# Executive Summary
The magazine Motor Trend is interested in exploring the relationship between a set of variables and miles per gallon (MPG). They are particularly interested in the following two questions:

* "Is an automatic or manual transmission better for MPG?”
* "Quantify the MPG difference between automatic and manual transmissions"

The best (i.e. model with the highest adjusted R^2) multivariate regression model was based on *number of cylinders*, *horsepower*, *car weight*, and *transmission type*; explaining 84.0% of variability in the data. These variables were selected using the Akaike Information Criterion (AIC). After validating this model the conclusion is that _cars with manual transmission are better for MPG and have on average a 1.80 higher mpg than cars with automatic transmission._

#Data
##Exploratory Data Analysis
This analysis is based on the mtcars dataset from the base r datasets-package. The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).
```{r}
data=mtcars
```
The appendix _(table 1 & plots 1-2)_ contains the results of basic data exploration. Main conclusions are: 

* the data does not seem to contain missing values and/or clear outliers. 
* there is a difference in fuel consumption (mpg) per Transmission Type.  _(see appendix, plot 2)_ 
* the distribution of fuel consuption (mpg) appears _normal_, this allows for fitting a linear model. _(see appendix, plot 1)_

##Preparing Data
The dataset is tidy and needs no cleaning. The variables am, cyl, vs, gear and carb will be converted to factors, the values 0 and 1 for transmssion type will be replaced by "Automatic" and "Manual".
```{r}
mtcars.clean<-mtcars
mtcars.clean$am <- as.factor(mtcars$am)
mtcars.clean$cyl <- as.factor(mtcars$cyl)
mtcars.clean$vs <- as.factor(mtcars$vs)
mtcars.clean$gear <- as.factor(mtcars$gear)
mtcars.clean$carb <- as.factor(mtcars$carb)
levels(mtcars.clean$am) <-c("Automatic", "Manual")
```
# Hypothesis testing
A simple boxplot suggests that there is  a difference in fuel consumption per transmission type. _(see appendix, plot 2)_  Running a Welch Two Sample t-test could confirm whether the difference is significant and whether we can reject our null-hypothesis that _no significant difference exists between cars with automatic transmission and cars with manual transmission_.
```{r}
t.test(mtcars.clean[mtcars.clean$am == "Automatic",]$mpg, 
       mtcars.clean[mtcars.clean$am == "Manual",]$mpg)
```
The low p-value (0.001374) indicates that the probability that this difference is accidental is very low; the difference is *significant* and the null-hypothesis can be rejected. For further quantifying the effect we need to apply linear regression.

# Regression on a single variable
Perform a simple regression of fuel consumtion (mpg) on transmission type (am):
```{r results='hide'}
fit1<-lm(mpg ~ am, data=mtcars.clean)
```
The coefficents of this model are:
```{r}
summary(fit1)$coef
```
The intercept of 17.15 is the mean mpg of cars with automatic transmission. The amManual estimate is the expected change in mpg from automatic transmission to manual transmission. The low p-value (0.000285) might tempt us to conclude that cars with manual transmission have - on average - a 7.25 higher mpg. However, the adjusted R^2 value of this model is quite low:
```{r}
summary(fit1)$adj.r.squared
```
This means that only 34% of variation can be explained by this regression model; this is not enough to quantify a possible effect.

# Multivariate Regression

## Model Selection
Creating a model using all variables does not single out any variable with a significant p-value.
```{r}
fit.mv.0 <- lm(formula = mpg ~ ., data = mtcars.clean)
summary(fit.mv.0)
```
It does however mark weight and horsepower as significant candidates. Therefore, this will be the first multivariate model we will evaluate:```lm(formula = mpg ~ wt + hp + am, data = mtcars.clean)```

Another approach is to select a model on the Akaike Information Criterion (AIC) using a stepwise algorithm:  ``` step(fit.mv.0,direction="both", k=2)```  This produces the second multivariate model we will evaluate:```lm(formula = mpg ~ cyl + hp + wt + am, data = mtcars.clean)```

## Model Comparison
The conclusion will be based on the model having the highest Adjusted R^2 value.

#### Model 1 (wt + hp + am)

The R^2 value of the model using weight, horsepower, and transmission type:

```{r}
fit.mv.1 <- lm(formula = mpg ~ wt + hp + am, data = mtcars.clean)
summary(fit.mv.1)$adj.r.squared
```

#### Model 2 (cyl + hp + wt + am)

The R^2 value of the model using nbr of cylinders, horsepower, weight, and transmission type:

```{r}
fit.mv.2 <- (lm(formula = mpg ~ cyl + hp + wt + am, data = mtcars.clean))
summary(fit.mv.2)$adj.r.squared
```
We will use model 2 that was selected using AIC because of its (slightly) higher Adjusted R^2, provided no anomalies pop up during model diagnosis.


## Model Diagnosis
The diagnostic plots _(see appendix, plot 3)_ confirm that 

* no pattern exists in the residuals vs. fitted plot (i.e. independence)
* the Q-Q plot produces a line (i.e. residuals are distributed normally)
* no pattern exists in the scale-location plot (i.e. constant variance)
* the data contains no influential outliers in the residuals vs. leverage plot.

# Appendix        
## Tables

### Table 1: Summary of MPG per transmission type
```{r}
by(mtcars$mpg, mtcars$am, summary)
```

## Plots

### Plot 1: Density plot MPG
```{r}
plot(density(mtcars$mpg))
```

### Plot 2: Boxplot MPG and Transmission Type
```{r warning=FALSE}
library(ggplot2)
ggplot(mtcars.clean, aes(x=am, y=mpg, fill=am)) + 
        geom_boxplot() + 
        geom_jitter() +  
        ylab("Miles per Gallon") + 
        xlab("Transmission Type") + 
        theme(legend.title=element_blank())
```

### Plot 3: Diagnostics
```{r}
par(mfrow = c(2, 2))
plot(fit.mv.2)
```
